Optimal Light Curve Attitude Inversion with Measurement Noise: Two Case Studies

### 2025-03-24 paper.tex
almost done

\usepackage{subcaption}
\usepackage{times}
\usepackage[numbers]{natbib}
\usepackage{graphicx}
\usepackage{bm} % for bold math
\usepackage{mathtools} % for \prescript
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{adjustbox}

@@

In this work, we apply a BFGS-based multi-start global attitude inversion algorithm to synthetic observations of an upper stage and real observations taken by the Purdue Optical Ground Station of a Delta I upper stage and the ECHOSTAR II satellite. Given a light curve with an estimate of the shape and reflective properties of the object, our procedure searches the space of initial orientations, angular velocities, and inertia tensors to find initial conditions that produce light curves with low error compared to the observed values. Our formulation inherently accounts for the aforementioned ambiguities, the physical constraints of torque-free rigid body motion, and the noise in the light curve.

@@

\begin{itemize}
  \item Add the constant term back into the objective function (in text and code) so we can start treating it as a real likelihood, useful for the future
  \item Mention the unknown solar panel rotation angle for ECHOSTAR
  \item Update perspective on some the case 1a angular velocity plot
  \item Update $n_\text{guess}$ values everywhere to reflect the scripts
  \item Update $\omega$ sampling scheme to go from $0.5-2$ as reflected in scripts
  \item Re-run all cases to ensure that values are up to date with recent changes, update where those statistics are reported in the text
  \item Try to get rid of big gaps where possible
\end{itemize}

@@

Methods for estimating the attitude state of uncontrolled space objects from light curves differ significantly depending on the presence of an initial guess. If a guess is available, many filter-based approaches have been developed to process new photometry to update the guess orientation and angular velocity. This class of methods includes single Kalman filters \cite{burton2021two, gagnon2024, wetterer2009}, and multi-filter multiple-model adaptive estimation (MMAE) schemes \cite{linares2014space, dianetti2020}. These approaches can be very effective if the initial state estimate is close to the truth, but have been reported to diverge even in cases when the initial guess guess is within $5^\circ$ of the ground truth orientation and $0.03^\circ/s^{-1}$ in angular velocity error \cite{gagnon2024} due to measurement nonlinearity and ambiguities.

If no initial attitude guess is available, as in most cases of passive debris observation, the inversion problem must be solved globally. Filtering approaches can be applied here too; particle \cite{linares2014particle, holzinger2014} and Bayesian Multi-Hypothesis filters \cite{burton2021two, cabrera2023} have shown promise in prior work, but still rely on high-quality priors as with the Kalman filters. In the absence of prior knowledge about the object's attitude state, simulated annealing \cite{gagnon2024, clark2020}, genetic algorithms \cite{gagnon2024, piergentili2017, clark2020}, and particle swarm optimizers (PSOs) \cite{clark2020, clark2022, burton2024journal, burton2024scitech, gagnon2024, gagnon2025} can be applied to search the solution space exhaustively.

While much work has performed full state attitude inversion on simulated light curve data \cite{burton2024journal, gagnon2024, robinson2025att}, studies using real data often focus on determining a constant spin rate and axis of rotation for the observed object.

@@

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve. We present inversion results for a simulated rocket body light curve where the ground truth is known, followed by two cases -- a derelict satellite and rocket body -- using real light curve data collected by the Purdue Optical Ground Station. A deeper treatment of our method and additional simulated light curve results can be found in \cite{robinson2025att}.

@@

\begin{equation} \label{eq:nll_loss}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\left[\ln\sigma_k + \frac{1}{2}\left(\frac{\vctr{S}_k - \hat{\vctr{S}}_k(\vctr{x})}{\sigma_k}\right)^2 \right].
 \end{equation}

If the reflectivity of the object's faces is not well-known, we rescale the estimated light curve mean $\hat{S}(\vctr{x})$ to have equal $\ell_2$ norm to the observations to resolve the inherent albedo-area ambiguity:

@@

Our multi-start solution method begins by creating $n_\text{guess}$ guesses scattered throughout the solution space -- possibly sampled from a prior distribution based on available knowledge. In the most basic case, the initial state guess is sampled for guess $i$ at iteration $k=0$:

@@

In this work, initial guesses for the state vector are sampled from a prior distribution, usually defined independently for the MRP, angular velocity, and inertia tensor ratios. If no information is available, a uniform distribution is sampled from for the first six components of the state. In particular, $\vctr{p}_{i,0}(0)$ is uniformly randomly sampled from orientation space, $\vctr{\omega}_{i,0}(0)$ is randomly sampled from a uniform distribution of magnitudes up to $\norm{\vctr{\omega}}_\text{max}$ -- the maximum angular rate at which to initialize guesses. We determine $\norm{\vctr{\omega}}_\text{max}$ based on the maximum power positive frequency in the observed light curve $f^\star$ in Hertz and an arbitrary safety factor $f_\omega \geq 1$, selected as $f_\omega = 1.25$ for this work:

@@

Uniformly sampling over angular velocity directions, we can express the default sampling strategy for an angular velocity guess as:

@@

\noindent
where $g_{1:3} \in \mathcal{N}(0,1)$ and $m=\mathcal{U}(0,\norm{\vctr{\omega}}_\text{max})$.

Even for a well-chosen bound $\norm{\vctr{\omega}}_\text{max}$, there may still be numerous high-quality solutions at significantly lower or higher angular velocities that possess equal $f^\star$ values due to object shape symmetries or harmonics in spin and precession rates, respectively. As a result, it is usually unwise to directly constrain the angular velocity to a given magnitude unless additional, unambiguous information is available.

Similarly for the MRP, we first sample a quaternion $\vctr{q}_i(0) = [q_x, q_y, q_z, q_w]^T$ uniformly over orientation space:

@@

\noindent
and project the quaternion into MRP space:

@@

Unlike the MRP and angular velocity, the default sampling method for inertia ratios initial guesses $\left(J_y / J_x\right)_{i,0}$ and $\left(J_z / J_x\right)_{i,0}$ draws from independent, identically distributed Gaussian prior distributions with $\sigma_J$ centered at estimated values derived from the known geometry of the object $\left(J_y / J_x\right)_\text{est}$ and $\left(J_z / J_x\right)_\text{est}$, such that:

@@

After computing each initial state guess, we run the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{broyden1970, fletcher1970, goldfarb1970, shanno1970} to optimize each guess to a nearby local minimum $\vctr{x}_{k,i}(0)$ in the objective function such that:

@@

Although all initial conditions will converge to some nearby local minimum, many of these minima will possess a low likelihood determined by $-\mathrm{exp}(f(\vctr{x}))$. As we do not generally know how large the geometric and material mismatch between the observed object and the assumed model is, we cannot put a strict bound on a likelihood level that constitutes a high-quality candidate solution. In this work, we identify candidate solutions by choosing a likelihood ratio tolerance $0 < \ell < 1$, such that all solutions $\vctr{x}_{i,k}(0)$ are candidates for the truth if:

@@

 In any particular illumination and observation condition, a fraction of the facet's area $\tilde{a}_i$ may be Sun- or observer-shadowed due to obstructing geometry before or after light reaches the surface element, respectively. The remainder of the facet's area $\bar{a}_i$ is unshadowed and contributes to the total flux reaching the observer from the object, such that the total area of each facet is $a_i = \tilde{a}_i + \bar{a}_i$. The geometry for an individual facet is displayed in Figure \ref{fig:facet_geom}.

@@

The unshadowed area $\bar{a}$ may be computed in several ways. For convex objects, $\bar{a}=a$ as self-shadowing is impossible. For highly nonconvex and detailed models composed of thousands of faces, $\bar{a}$ can be approximated on a pixel grid using shadow mapping \cite{robinson2022}. Self-shadowing can be efficiently solved semi-analytically for lower-fidelity approximations of real space object geometry by computing the mutual intersections of the polygons $P_k$ of other faces whose projections overlap with the face in question. Up to a maximum intersection depth $d$, selected for computation time, the unshadowed area can be computed via:

@@

As the objects we investigate in this work are well-approximated with relatively simple non-convex meshes, we use the semi-analytic method to efficiently compute unshadowed areas via the Sutherland-Hodgeman algorithm \cite{sutherland1974} for each polygon intersection. Further explanation of the polygon clipping procedure for computing the total shadowed area and accelerating this computation during inversion with the so-called shadow cache is included here \textbf{TODO: should this be a `here'? wording this felt weird if we're avoiding saying like `the authors'"} \cite{robinson2025att}.

@@

The output of Equation \ref{eq:lc_norm} is a light curve, but as many of the noise characteristics of the signal are defined in the image sensor's native unit of ADU, we convert the light curve into the same units via:

@@

The sensor noise is approximated by the independent combination of Poisson dark current $\lambda_\text{dark}$ and readout noise $\sigma_\text{read}^2$ \cite{krag2003}:

@@

In order to make this simulated case realistic, we use the presented measurement noise models and introduce a mismatch between the ground truth object used to generate the measurements and the model used in inversion. We produce a noisy, observed light curve for a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown on the right of Figure \ref{fig:rb_models}.

@@

Table \ref{tb:case1_in} describes the simulated observation scenario, placing the rocket body in the orbit of an SL-12 upper stage in geosynchronous orbit, observed by the Purdue Optical Ground Station. Relevant geographical and optical properties of this ground station are provided in the Appendix in Table \ref{tb:tele_info}. Table \ref{tb:synth_matprops} lists the material properties used in this test case.

\begin{table}[H]
  \centering
  \caption{Simulated observation characteristics}
  \vspace*{6pt}
  \begin{tabular}{|l|l|}
  \hline
  \textbf{Variable} & \textbf{Value} \\ \hline
 Target COSPAR ID & 1990-054D \\ \hline
 First obs.\ (UTC) & Mar 8, 2025 02:00:00.000 \\ \hline
 Light curve duration & $5$ minutes \\ \hline
 Observations & $100$ \\ \hline
 Integration time & $0.5$ seconds \\ \hline
  \end{tabular}
  \label{tb:case1_in}
\end{table}

\begin{table}[H]
  \centering
  \caption{Reflection properties of materials used in the synthetic data test case. The white paint parameters are estimated qualitatively}
  \vspace*{6pt}
  \begin{tabular}{|l|l|l|l|}
  \hline
  \textbf{Material} & $C_d$ & $C_s$ & $n$ \\ \hline
 Aluminum \cite{fankhauser2023} & $0.34$ & $0.40$ & $8.9$ \\ \hline
 White paint & $0.9$ & $0.1$ & $1$ \\ \hline
  \end{tabular}
  \label{tb:synth_matprops}
\end{table}

@@

\begin{table}[H]
  \centering
  \caption{True object attitude profile for all test cases}
  \vspace*{6pt}
  \begin{tabular}{|l|l|}
  \hline
  \textbf{Variable} & \textbf{Value} \\ \hline
 Initial $\vctr{p}(0)$ & $-\frac{1}{3} \begin{bmatrix} 1 & 1 & 1 \end{bmatrix}^T$ \\ \hline
 True $\vctr{\omega}(0)$ ($\text{rad}\cdot\text{s}^{-1}$) & $\vctr{\omega}(0) = \begin{bmatrix} 0.03 & 0.06 & 0.03 \end{bmatrix}^T$ \\ \hline
 Inertia tensor ratios & $J_y / J_x = 1, \: J_z / J_x = 0.25$ \\ \hline
 External torque $\vctr{M}$ & $\begin{bmatrix} 0 & 0 & 0 \end{bmatrix}^T$ \\ \hline
  \end{tabular}
  \label{tb:synth_att}
\end{table}
\FloatBarrier

@@

Two distinct inversion cases are computed. Case 1a presents inversion results given no attitude information besides an estimate of the inertia tensor ratios derived from the known geometry, while Case 1b presents results given good initial guesses for orientation and angular velocity. Tables \ref{tb:case1a_prior} and \ref{tb:case1b_prior} summarize the initial condition sampling strategies employed for each case.

For both cases, a total of $n_\text{guess}=10^5$ initial state guesses are created via the sampling schemes described their respective tables. BFGS minimization is then performed for each initial guess to locally solve the optimization problem given by Equation \ref{eq:opt_problem}.

\begin{table}[H]
  \centering
  \caption{Initial condition sampling parameters used for inversion in Case 1a}
  \vspace*{6pt}
  \begin{tabular}{|l|l|}
  \hline
  \textbf{Variable} & \textbf{Value} \\ \hline
  $\vctr{p}$ sampling & Uniform, via Equation \ref{eq:mrp_sampler} \\ \hline
  $\vctr{\omega}$ sampling & Uniform, via Equation \ref{eq:omega_sampler} \\ \hline
  $J$ sampling & Gaussian, via Equation \ref{eq:omega_sampler} \\ \hline
  $\sigma_J$ & $0.05$ $[\text{nondim}]$ \\ \hline
  $\norm{\vctr{\omega}}_\text{max}$ & $5.26$ $[\text{deg}\cdot \text{s}^{-1}]$ \\ \hline
\end{tabular}
  \label{tb:case1a_prior}
\end{table}

\begin{table}[H]
  \centering
  \caption{Informative prior parameters used for inversion in Case 1b. The standard deviations $\sigma_\omega$ and $\sigma_J$ are for spherically-symmetric Gaussians with centered on the ground truth. MRPs are sampled by rotating by $\theta\sim \mathcal{N}(0,\sigma_p)$ degrees away from the ground truth orientation around an axis uniformly sampled on the sphere.}
  \vspace*{6pt}
  \begin{tabular}{|l|l|}
  \hline
  \textbf{Variable} & \textbf{Value} \\ \hline
  $\vctr{p}$ sampling & Gaussian, $\sigma_\omega = 0.05$ $[\text{deg}\cdot\text{s}^{-1}]$ \\ \hline
  $\vctr{\omega}$ sampling & Gaussian, $\sigma_p=20$ $[\text{deg}]$ \\ \hline
  $J$ sampling & Gaussian, $\sigma_J = 0.05$ $[\text{nondim}]$ \\ \hline
\end{tabular}
  \label{tb:case1b_prior}
\end{table}

Solutions are then classified as candidates for the ground truth with $\ell = 1/10$ via Equation \ref{eq:ell_selection_criteria}. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.

@@

Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} display the candidate MRP and angular velocity of the candidate solutions determined via the procedure in Section \ref{sec:candidate_sols}. In the left Case 1a figures for orientation and angular velocity, we notice that the ground truth lies squarely on a surface containing numerous candidate solution. This indicates that the local minima have correctly converged to the non-singular subspace of states that approximate the observations well for the assumed shape model. In angular velocity space, shown in Figure \ref{fig:omega_sols1}, most solutions fall onto the surface of two open-ended cylinders -- one with half the radius of the larger cylinder containing the truth. Similarly in orientation space, shown in Figure \ref{fig:sigma_sols1}, the solutions fall on two distinct manifolds, one containing the truth -- likely resulting from a reflection of the object's body frame about the bisector between the Sun and observer vectors \cite{marto2024, burton2024journal}. These additional families of solutions highlight the reality that a wide variety of attitude profiles may produce nearly identical observations.

The Case 1b results on the right within Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} show that the majority of the initial guesses sampled close to the truth have converged to a local subset of the solution manifolds identified by Case 1a. We should not expect these solutions to have shrunk down exponentially towards the ground truth -- the mismatch in the assumed shape model and the measurement noise inherently limits the minimum size of the feasible solution space.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body.png}
    \caption{}
    \label{fig:sigma_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body Prior.png}
    \caption{}
    \label{fig:sigma_sols1b}
  \end{subfigure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body.png}
    \caption{}
    \label{fig:omega_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body Prior.png}
    \caption{}
    \label{fig:omega_sols1b}
  \end{subfigure}
  \caption{Candidate solution initial body-frame angular velocity vectors for Cases 1a (left) and 1b (right), with the ground truth highlighted in red. TODO: change the perspective on the angular velocities}
  \label{fig:omega_sols1}
\end{figure}

Figure \ref{fig:i_sols1} displays the inertia tensor solutions for Cases 1a and 1b. Notably, the majority of candidate solutions remain within a few standard deviations of the mean of their initial distribution, with the remaining minority exploring directions that maintain the same precession rate. When a good initial guess for the other attitude state components is available in Case 1b, the inertia tensor solutions are more well-clustered, although due to the low observability of the rocket body's spin rate in the observed light curve, there are still ambiguous solutions with higher or lower $J_z / J_x$ values.

>>>>

\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{times}
\usepackage[numbers]{natbib}
\usepackage{graphicx}
\usepackage{bm} % for bold math
\usepackage{mathtools} % for \prescript
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{adjustbox}

\newcommand{\grule}[0]{\arrayrulecolor{darkgray}\cmidrule(){1-2}}
\newcommand{\brule}[0]{\arrayrulecolor{black} \bottomrule}

@@

In this work, we apply a BFGS-based multi-start global attitude inversion algorithm to efficiently solve the attitude inversion problem globally. Results for synthetic observations are presented for a generic upper stage. Results for real observations taken by the Purdue Optical Ground Station of Delta I upper stage and the ECHOSTAR II satellite are also presented. Given a light curve with an estimate of the shape and reflective properties of the object, our procedure searches the space of initial orientations, angular velocities, and inertia tensors to find initial conditions that produce light curves with low error compared to the observed values. Our formulation inherently accounts for the aforementioned ambiguities, the physical constraints of torque-free rigid body motion, and the noise in the light curve.

@@

@@

Methods for estimating the attitude state of uncontrolled space objects from light curves differ significantly depending on the presence of an initial guess. If prior knowledge is available, many filter-based approaches have been developed to process new photometry to update the guess orientation and angular velocity. This class of methods includes single Kalman filters \cite{burton2021two, gagnon2024, wetterer2009}, and multi-filter multiple-model adaptive estimation (MMAE) schemes \cite{linares2014space, dianetti2020}. These approaches can be very effective if the initial state estimate is close to the truth, but have been reported to diverge even in cases when the initial guess is within $5^\circ$ of the ground truth orientation and $0.03^\circ/s^{-1}$ in angular velocity error \cite{gagnon2024} due to measurement nonlinearity and ambiguities.

If no initial attitude guess is available, as in most cases of passive debris observation, the inversion problem must be solved globally. Filtering approaches can be applied here too; particle \cite{linares2014particle, holzinger2014} and Bayesian Multi-Hypothesis filters \cite{burton2021two, cabrera2023} have shown promise in prior work, but still rely on high-quality priors as with the single-hypothesis Kalman filters. In the absence of any prior knowledge about the object's attitude state, simulated annealing \cite{gagnon2024, clark2020}, genetic algorithms \cite{gagnon2024, piergentili2017, clark2020}, and particle swarm optimizers (PSOs) \cite{clark2020, clark2022, burton2024journal, burton2024scitech, gagnon2024, gagnon2025} can be applied to search the solution space exhaustively.

While much work has performed full state attitude inversion on simulated light curve data \cite{burton2024journal, gagnon2024, robinson2025att}, studies using real data often focus on estimating an assumed constant spin rate and axis of rotation for the observed object. These assumptions simplifies the estimation process, but can lead to misleading conclusions when the observed object has a complex rotation state.

@@

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve. We present inversion results for a simulated rocket body light curve where the ground truth is available to demonstrate the efficacy of the method, followed by results for real light curve data collected by the Purdue Optical Ground Station for a derelict ECHOSTAR satellite and Delta I rocket body. A deeper treatment of our method and additional simulated light curve results can be found in \cite{robinson2025att}.

@@

\begin{equation} \label{eq:nll_loss}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\left[\frac{1}{2} \ln2\pi + \ln\sigma_k + \frac{1}{2}\left(\frac{\vctr{S}_k - \hat{\vctr{S}}_k(\vctr{x})}{\sigma_k}\right)^2 \right].
 \end{equation}

Note that the constant $0.5\ln2\pi$ could be removed from the objective function, but it is retained such that $\text{exp}\left(-f(\vctr{x})\right)$ is a true likelihood. If the reflectivity of the object's faces is not well-known, we rescale the estimated light curve mean $\hat{S}(\vctr{x})$ to have equal $\ell_2$ norm to the observations to resolve the inherent albedo-area ambiguity:

@@

Our multi-start solution method begins by creating $n_\text{sample}$ sampled states scattered throughout the solution space -- possibly sampled from a prior distribution based on available knowledge. In the most basic case, the initial states are drawn for sample $i$ at iteration $k=0$:

@@

In this work, initial state vectors are sampled from a prior distribution, usually defined independently for the MRP, angular velocity, and inertia tensor ratios. If no information is available, as is usually the case, a uniform distribution is sampled from for the first six components of the state. In particular, $\vctr{p}_{i,0}(0)$ is uniformly randomly sampled from orientation space, $\vctr{\omega}_{i,0}(0)$ is randomly sampled from a uniform distribution of magnitudes up to $\norm{\vctr{\omega}}_\text{max}$ -- the maximum angular rate at which to initialize state samples. We determine $\norm{\vctr{\omega}}_\text{max}$ based on the maximum power positive frequency in the observed light curve $f^\star$ in Hertz and an arbitrary safety factor $f_\omega \geq 1$, selected as $f_\omega = 1.25$ for this work:

@@

Uniformly sampling over angular velocity directions, we can express the default sampling strategy for an angular velocity state as:

@@

\noindent
where $g_{1:3} \in \mathcal{N}(0,1)$ and $m=\mathcal{U}(c_1 \norm{\vctr{\omega}}_\text{max}, c_2 \norm{\vctr{\omega}}_\text{max})$. The constants $c_1<1$ and $c_2>1$ are chosen to correct for the fact that even for a well-chosen bound $\norm{\vctr{\omega}}_\text{max}$, there can still be numerous high-quality solutions at significantly lower or higher angular velocities that possess equal $f^\star$ values. This is a result of object shape symmetries or harmonics in spin and precession rates. We select $c_1=0.5$, $c_2=2$ for this work.

For the MRP, we first sample a quaternion $\vctr{q}_i(0) = [q_x, q_y, q_z, q_w]^T$ uniformly over orientation space:

@@

\noindent
where $g_{1:4} \in \mathcal{N}(0,1)$, and project the quaternion into MRP space:

@@

Unlike the MRP and angular velocity, the default sampling method for inertia ratios initial samples $\left(J_y / J_x\right)_{i,0}$ and $\left(J_z / J_x\right)_{i,0}$ draws from independent, identically distributed Gaussian prior distributions with $\sigma_J$ centered at estimated values derived from the known geometry of the object $\left(J_y / J_x\right)_\text{est}$ and $\left(J_z / J_x\right)_\text{est}$, such that:

@@

After sampling each initial state, we run the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm \cite{broyden1970, fletcher1970, goldfarb1970, shanno1970} to optimize each sample to a nearby local minimum $\vctr{x}_{k,i}(0)$ in the objective function such that:

@@

Although all initial conditions will converge to some nearby local minimum, many of these minima will possess a low likelihood determined by $-\mathrm{exp}(f(\vctr{x}))$ where $f(\vctr{x})$ is the objective function in Equation \ref{eq:nll_loss}. As we do not generally know how large the geometric and material mismatch between the observed object and the assumed model is, we cannot put a strict bound on a likelihood level that constitutes a high-quality candidate solution. In this work, we identify candidate solutions by choosing a likelihood ratio tolerance $0 < \ell < 1$, such that all solutions $\vctr{x}_{i,k}(0)$ are candidates for the truth if:

@@

 In any particular illumination and observation condition, a fraction of the facet's area $\tilde{a}_i$ might be Sun or observer-shadowed due to obstructing geometry before or after light reaches the surface element, respectively. The remainder of the facet's area $\bar{a}_i$ is unshadowed and contributes to the total flux reaching the observer from the object, such that the total area of each facet is $a_i = \tilde{a}_i + \bar{a}_i$. The geometry for an individual facet is displayed in Figure \ref{fig:facet_geom}.

@@

The unshadowed area $\bar{a}$ can be computed in several ways. For convex objects, $\bar{a}=a$ as self-shadowing is impossible. For highly nonconvex and detailed models composed of thousands of faces, $\bar{a}$ can be approximated on a pixel grid using shadow mapping \cite{robinson2022}. Self-shadowing can be efficiently solved semi-analytically for lower-fidelity approximations of real space object geometry by computing the mutual intersections of the polygons $P_k$ of other faces whose projections overlap with the face in question. Up to a maximum intersection depth $d$, selected for computation time, the unshadowed area can be computed via:

@@

As the objects we investigate in this work are well-approximated with relatively simple non-convex meshes, we use the semi-analytic method to efficiently compute unshadowed areas via the Sutherland-Hodgeman algorithm \cite{sutherland1974} for each polygon intersection. Further explanation of the polygon clipping procedure for computing the total shadowed area, and accelerating this computation during inversion with the so-called shadow cache is available here \cite{robinson2025att}.

@@

As many of the noise characteristics of the signal are defined in the image sensor's native unit of ADU, we use $f_p(\unitv{s}_k, \unitv{o}_k)$ to compute the light curve in ADU via:

@@

The sensor noise is approximated by the independent combination of Poisson dark current $\lambda_\text{dark}$ and zero-mean Gaussian readout noise $\sigma_\text{read}^2$ \cite{krag2003}:

@@

In order to make this simulated case realistic, we use the presented measurement noise models and introduce a mismatch between the ground truth object used to generate the measurements and the model used in inversion. The selected object is a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown on the right of Figure \ref{fig:rb_models}

@@

Table \ref{tb:case1_in} describes the simulated observation scenario used to produce the noisy light curve observations, placing the rocket body in the orbit of an SL-12 upper stage in geosynchronous orbit, observed by the Purdue Optical Ground Station. Relevant geographical and optical properties of this ground station are provided in the Appendix in Table \ref{tb:tele_info}. Table \ref{tb:synth_matprops} lists the material properties used in this test case.

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Simulated observation characteristics}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    Target COSPAR ID & 1990-054D \\ \grule
    First obs.\ (UTC) & Mar 8, 2025 02:00:00.000 \\ \grule
    Light curve duration & $5$ minutes \\ \grule
    Observations & $100$ \\ \grule
    Integration time & $0.5$ seconds \\ \brule
  \end{tabular}
  \label{tb:case1_in}
\end{table}

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Reflection properties of materials used in the synthetic data test case. The white paint parameters are estimated qualitatively.}
  \vspace*{6pt}
  \begin{tabular}{@{} l l l l @{}}
    \toprule
    Material & $C_d$ & $C_s$ & $n$ \\ \midrule
    Aluminum \cite{fankhauser2023} & $0.34$ & $0.40$ & $8.9$ \\ \arrayrulecolor{darkgray}\cmidrule(){1-4}
    White paint & $0.9$ & $0.1$ & $1$ \\ \brule
  \end{tabular}
  \label{tb:synth_matprops}
\end{table}

@@

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{True object attitude profile for all test cases}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    Initial $\vctr{p}(0)$ & $-\frac{1}{3} \begin{bmatrix} 1 & 1 & 1 \end{bmatrix}^T$ \\ \grule
    True $\vctr{\omega}(0)$ ($\text{rad}\cdot\text{s}^{-1}$) & $\vctr{\omega}(0) = \begin{bmatrix} 0.03 & 0.06 & 0.03 \end{bmatrix}^T$ \\ \grule
    Inertia tensor ratios & $J_y / J_x = 1, \: J_z / J_x = 0.25$ \\ \grule
    External torque $\vctr{M}$ & $\begin{bmatrix} 0 & 0 & 0 \end{bmatrix}^T$ \\ \brule
  \end{tabular}
  \label{tb:synth_att}
\end{table}
\FloatBarrier

@@

Two distinct inversion cases are computed. Case 1a presents inversion results given no attitude information besides an estimate of the inertia tensor ratios derived from the known geometry, while Case 1b presents results given good prior attitude state distributions. Tables \ref{tb:case1a_prior} and \ref{tb:case1b_prior} summarize the initial condition sampling strategies employed for each case.

For both cases, a total of $n_\text{sample}=10^5$ initial state samplees are created via the sampling schemes described their respective tables. BFGS minimization is then performed for each initial state sample to locally solve the optimization problem given by Equation \ref{eq:opt_problem}.

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Initial condition sampling parameters used for inversion in Case 1a}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    $\vctr{p}$ sampling & Uniform, via Equation \ref{eq:mrp_sampler} \\ \grule
    $\vctr{\omega}$ sampling & Uniform, via Equation \ref{eq:omega_sampler} \\ \grule
    $J$ sampling & Gaussian, via Equation \ref{eq:omega_sampler} \\ \grule
    $\sigma_J$ & $0.1$ $[\text{nondim}]$ \\ \grule
    $\norm{\vctr{\omega}}_\text{max}$ & $5.26$ $[\text{deg}\cdot \text{s}^{-1}]$ \\ \brule
  \end{tabular}
  \label{tb:case1a_prior}
\end{table}

\begin{table}[H]
  \centering
  \renewcommand{\arraystretch}{1.3} % Increase row height for better readability
  \caption{Informative prior parameters used for inversion in Case 1b. The standard deviations $\sigma_\omega$ and $\sigma_J$ are for spherically-symmetric Gaussians centered on the ground truth. MRPs are sampled by rotating by $\theta\sim \mathcal{N}(0,\sigma_p)$ degrees away from the ground truth orientation around an axis uniformly sampled on the sphere.}
  \vspace*{6pt}
  \begin{tabular}{@{} l l @{}}
    \toprule
    Variable & Value \\ \midrule
    $\vctr{p}$ sampling & Gaussian, $\sigma_p=20$ $[\text{deg}]$ \\ \grule
    $\vctr{\omega}$ sampling & Gaussian, $\sigma_\omega = 0.1$ $[\text{deg}\cdot\text{s}^{-1}]$ \\ \grule
    $J$ sampling & Gaussian, $\sigma_J = 0.05$ $[\text{nondim}]$ \\ \brule
  \end{tabular}
  \label{tb:case1b_prior}
\end{table}

Solutions are then classified as candidates for the ground truth with $\ell = 1/2$ via Equation \ref{eq:ell_selection_criteria}. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.

@@

Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} display the candidate MRP and angular velocity of the candidate solutions determined via the procedure in Section \ref{sec:candidate_sols}. In the left Case 1a figures for orientation and angular velocity, we notice that the ground truth in red lies squarely on a surface containing numerous candidate solutions. This indicates that the local minima have correctly converged to the non-singular subspace of states that approximate the observations well for the assumed shape model. In angular velocity space, shown in Figure \ref{fig:omega_sols1}, most solutions fall onto the surface of two open-ended cylinders -- one with half the radius of the larger cylinder containing the truth. Similarly in orientation space, shown in Figure \ref{fig:sigma_sols1}, the solutions fall on two distinct manifolds, one containing the truth -- likely resulting from a reflection of the object's body frame about the bisector between the Sun and observer vectors \cite{marto2024, burton2024journal}. These additional families of solutions highlight the reality that a wide variety of attitude profiles can produce nearly identical observations.

The Case 1b results on the right within Figures \ref{fig:sigma_sols1} and \ref{fig:omega_sols1} show that the majority of the initial states sampled close to the truth have converged to a local subset of the solution manifolds identified by Case 1a. We should not expect these solutions to have shrunk down exponentially towards the ground truth -- the mismatch in the assumed shape model and the measurement noise inherently limits the minimum size of the feasible solution space.

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body.png}
    \caption{}
    \label{fig:sigma_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{sigma_solsRocket Body Prior_pin.png}
    \caption{}
    \label{fig:sigma_sols1b}
  \end{subfigure}

@@

\begin{figure}[H]
  \centering
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body_pin.png}
    \caption{}
    \label{fig:omega_sols1a}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.23\textwidth}
    \centering
    \includegraphics[width=\textwidth]{omega_sols_Rocket Body Prior_pin.png}
    \caption{}
    \label{fig:omega_sols1b}
  \end{subfigure}
  \caption{Candidate solution initial body-frame angular velocity vectors for Cases 1a (left) and 1b (right), with the ground truth highlighted in red. TODO: change the perspective on the angular velocities}
  \label{fig:omega_sols1}
\end{figure}

Figure \ref{fig:i_sols1} displays the inertia tensor solutions for Cases 1a and 1b. Notably, the majority of candidate solutions remain within a few standard deviations of the mean of their initial distribution, with the remaining minority exploring directions that maintain the same precession rate. When a good prior for the other attitude state components is available in Case 1b, the inertia tensor solutions are more well-clustered, although due to the low observability of the rocket body's spin rate in the observed light curve, there are still ambiguous solutions with higher or lower $J_z / J_x$ values.
